---
title: "BenHuynhEDA"
output: html_document
date: "2025-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(GGally)   
#install.packages("ggforce")
library(ggforce)
library(corrplot)

bat <- read_csv("C:/Users/BenAH/Downloads/stat170proj/team_stan_bat.csv")
field <- read_csv("team_field.csv")
pitch <- read_csv("team_stan_pitch.csv")
pos <- read_csv("wins_above_avg_pos.csv")


head(bat)
head(field)
head(pitch)
head(pos)

```





```{r}

data <- read_csv("combined_teams.csv")

head(data)

```


1. Set Seed and Split Test/Train

```{r}
set.seed(123)

n <- nrow(data)
train_index <- sample(1:n, size = 0.7 * n)
train_data <- data[train_index, ]
test_data  <- data[-train_index, ]
```



```{r}
y_var <- "W" 

correlations <- train_data %>%
  select(where(is.numeric)) %>%               
  summarise(across(-all_of(y_var),             
                   ~ cor(.x, .data[[y_var]], use = "complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))           

correlations
```


2. Data Description

Size: 21 x 99
Response Variable: Y = W or Wins (#of games a team won in the season)

```{r}

dim(train_data)

```
Source : https://www.baseball-reference.com/leagues/majors/2024.shtml

Context: Statistics Gatheredf from the 30 MLB teams during the 2024 Season

(All taken from MLB Website)
#Bat -- Number of Players used in Games
BatAge -- Batters’ average age
Weighted by AB + Games Played
R/G -- Runs Scored Per Game
G -- Games Played or Pitched
PA -- Plate Appearances
When available, we use actual plate appearances from play-by-play game accounts
Otherwise estimated using AB + BB + HBP + SF + SH,
which excludes catcher interferences.
When this color click for a summary of each PA.
AB -- At Bats
R -- Runs Scored/Allowed
H -- Hits/Hits Allowed
HR -- Home Runs Hit/Allowed
RBI -- Runs Batted In
SB -- Stolen Bases
CS -- Caught Stealing
BB -- Bases on Balls/Walks
SO -- Strikeouts
BA -- Hits/At Bats
OBP -- (H + BB + HBP)/(At Bats + BB + HBP + SF)
For recent years, leaders need 3.1 PA
per team game played
SLG -- Total Bases/At Bats or
(1B + 2*2B + 3*3B + 4*HR)/AB
For recent years, leaders need 3.1 PA
per team game played
OPS -- On-Base + Slugging Percentages
For recent years, leaders need 3.1 PA
per team game played
OPS+ -- OPS+
100*[OBP/lg OBP + SLG/lg SLG - 1]
Adjusted to the player’s ballpark(s)
TB -- Total Bases
Singles + 2 x Doubles + 3 x Triples + 4 x Home Runs.
GDP -- Double Plays Grounded Into
Only includes standard 6-4-3, 4-3, etc. double plays.
For gamelogs only in seasons we have play-by-play, we include triple plays as well.
HBP -- Times Hit by a Pitch.
SH -- Sacrifice Hits (Sacrifice Bunts)
SF -- Sacrifice Flies
First tracked in 1954.
IBB -- Intentional Bases on Balls
First tracked in 1955.
LOB -- Runners Left On Base
#P -- Number of Pitchers used in Games
PAge -- Pitchers’ average age
Weighted by 3*GS + G + SV
RA/G -- Runs Allowed Per Game
W -- Wins
L -- Losses
W-L% -- Win-Loss Percentage
W / (W + L)
ERA -- 9 * ER / IP
For recent years, leaders need 1 IP
per team game played.
Bold indicates lowest ERA using current stats
Gold means awarded ERA title at end of year.
G -- Games Played or Pitched
GS -- Games Started
GF -- Games Finished
CG -- Complete Game
tSho -- Shutouts by a team
No runs allowed in a game by one or more pitchers.
cSho -- Shutouts
No runs allowed and a complete game.
SV -- Saves
IP -- Innings Pitched
H -- Hits/Hits Allowed
R -- Runs Scored/Allowed
ER -- Earned Runs Allowed
HR -- Home Runs Hit/Allowed
BB -- Bases on Balls/Walks
IBB -- Intentional Bases on Balls
First tracked in 1955.
SO -- Strikeouts
HBP -- Times Hit by a Pitch.
BK -- Balks
WP -- Wild Pitches
BF -- Batters Faced
ERA+ -- ERA+
100*[lgERA/ERA]
Adjusted to the player’s ballpark(s).
FIP -- Fielding Independent Pitching
this stat measures a pitcher's effectiveness at preventing HR, BB, HBP and causing SO
(13*HR + 3*(BB+HBP) - 2*SO)/IP + Constantlg
The constant is set so that each season major-league average FIP is the same as the major-league avg ERA
WHIP -- (BB + H)/IP
For recent years, leaders need 1 IP
per team game played
H9 -- 9 x H / IP
For recent years, leaders need 1 IP
per team game played
HR9 -- 9 x HR / IP
For recent years, leaders need 1 IP
per team game played
BB9 -- 9 x BB / IP
For recent years, leaders need 1 IP
per team game played
SO9 -- 9 x SO / IP
For recent years, leaders need 1 IP
per team game played
SO/W -- SO/W or SO/BB
For recent years, pitching leaders need 1 IP
per team game played.
No batting leaders computed.
LOB -- Runners Left On Base
ALL P -- All Positions
SP -- Start Pitcher
RP -- Reflief Pitcher
Non-P -- non-Pitcher
C -- Catcher
1B -- 1st Base
2B -- 2nd Base
3b -- 3rd Base
SS -- ShortStop
LF -- Left Field
CF -- Center Field
RF -- Right Field
OF(All) - Outfield(ALL)
DH - Designated Hitter
PH - Pinch Hitter
#Fld -- Number of Players used as Fielders
RA/G -- Runs Allowed Per Game
DefEff -- Defensive Efficiency
Percentage of balls in play converted into outs
This is an estimate based on team defensive and pitching stats.
We utilize two estimates of plays made.
One using innings pitched, strikeouts, double plays and outfield assists.
And the other with batters faced, strikeouts, hits allowed, walks allowed, hbp, and .71 * errors committed (avg percent of errors that result in an ROE)
Total plays available are plays made + hits allowed - home runs + error committed estimate.
G -- Games Played or Pitched
GS -- Games Started
CG -- Complete Game
Inn -- Innings Played in Field
Ch -- Defensive Chances
Putouts + Assists + Errors
PO -- Putouts
A -- Assists
E -- Errors Committed
DP -- Double Plays Turned
Fld% -- Fielding Percentage
(Putouts + Assists) / (Putouts + Assists + Errors)
Rtot -- Total Zone Total Fielding Runs Above Avg
The number of runs above or below average the player was worth based on the number of plays made.
This number combines the Rtz, Rdp, Rof, Rcatch numbers into a total defensive contribution.
See the glossary section for a more complete explanation.
Provided by BaseballProjection.com
Rtot/yr -- Total Zone Total Fielding Runs Above Avg per 1,200 Inn
The number of runs above or below average the fielder was worth per 1,200 Innings (approx 135 games).
This number combines the Rtz, Rdp, Rof, Rcatch numbers into a total defensive contribution.
See the glossary section for a more complete explanation.
Provided by BaseballProjection.com
Rdrs -- BIS Defensive Runs Saved Above Avg
The number of runs above or below average the team was worth based on their overall defense.
This number combines the individual players' defensive runs saved as well as team-level runs saved from shifting and positioning into a total team defensive contribution.
Provided by Baseball Info Solutions
Rdrs/yr -- BIS Defensive Runs Saved Above Avg per 1,200 Inn
The number of runs above or below average the team's fielders were worth per 1,200 Innings (approx 135 games).
This number combines the Rpm, Rdp, and Rgood numbers into a total defensive contribution.
For pitchers, this is set to 200 Innings.
This does not include any team-level runs saved from positioning or shifting.
Provided by Baseball Info Solutions
Rgood -- BIS Good Plays/Misplays Runs Above Avg



3. Descriptive Statistics

```{r}

#install.packages("skimr")
library(skimr)

skim(train_data)


```

#. variabe: shape w/, skew if any, ~mean, f~ median

1. approx norm, slight right skew, 55.24, 55
2. norm, 27.83, 27.8
3. not norm, slight right skew, 4.38, 4.28
4. norm, 6085, 6067
5. norm, 5456, 5460
6. not norm, right skew, 1321, 1318
7. not norm, right skew, 259, 257
8. not norm, right skew, 259, 257
9. not norm, right skew, 23, 23
10. not norm, right skew, 183, 179
11. not norm, right skew, 677, 661
12. not norm, right skew 677, 661
13. not norm, righ skew, 32, 26
14. norm, 499, 490
15. not normal, right skew, 1398, 1382
16. norm, 0.242, 0.242
17. norm, 0.311, 0.309
18. approx norm, right skew, 0.398, 0.393
19. approx norm, right skew, 0.710, 0.703
20. norm, 100.19, 100
21. not norm, right skew, 2174, 2139
22. not norm, right skew, 105, 105
23. not norm, right skew, 69, 64
24. not norm, right skew, 14, 14
25. norm, 42, 43
26. not norm, right skew, 18, 15
27. norm, 1083, 1099
28. norm, 55, 55
29. approx norm, right skew, 4.4, 4.31
30. norm, 0.698, 0.702
31. norm, 1076, 1080
32. norm, 12940, 12936
33. norm, 5779, 5771
34. norm, 4314, 4312
35. norm, 1378, 1366
36. approx norm, slight right skew, 88, 87
37. not norm, right skew, 122, 123
38. norm, 0.985, 0.985
39. not norm, left skew, -0.62, 4
40. not norm, still symmetric, -0.048, 0
41. not norm, slight left skew, -1.43, -3
42. not norm, still symmetric, 0, 0
43. not norm, slight left skew, -0.81, -1
44. norm, 33, 32
45. norm, 28.9, 28.9
46. norm, 81.2, 82
47, approx norm, right skew, 4.07, 4.04
48. not norm, right skew, 10.3, 11
49. not norm, right skew, 0.43, 0
50. approx norm, right skew, 43, 40
51. norm, 1437, 1437
52. norm, 1324, 2303
53. norm, 713, 699
54. norm, 650, 648
55. norm, 182, 181,
56. norm, 494, 492
57. nor norm right skew, 15, 14
58. norm, 1371, 1371
59. not norm, right skew, 68, 66
60. not norm, right skew, 68, 66
61. approx norm, right skew, 48, 46
62. norm, 6080, 6078
63. norm, 101, 101
64. norm, 4.07, 4.04
65. norm, 1.26, 1.25
66. norm, 8.29, 8.2
67. norm, 1.13, 1.1
68. norm, 3.1, 3.1
69. norm, 8.59. 8.6
70. norm 2.81, 2.78
71. norm, 1074, 1080
72. not norm, right skew, 16.6, 17
73. not norm, left skew, -0.46, 0.4
74. not norm, left skew, -0.27, 0
75. approx norm, right skew, 0.53, 1
76. not norm, left skew, -1.03, -0.9
77. not norm, left skew, -0.31, -0.6
78. not norm, left skew, -0.21, -0.2
79. not norm, left skew, -1.03, -1
80/ norm, -0.2, -0.3
81. norm, 0.048, 0.2
82. not norm, right skew, 1.53, 1.6
83. norm, -0.18, -0.1
84. approx norm, right skew, 0.47, -0.2
85. norm, -0.16, -0.3
86. not norm, right skew, 0.13, -0.8
87. norm, 0.009, -0.3
88. not norm, left skew, -0.6, -0.6


4. Visual Explanation

```{r}




long_data <- train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")


n_vars <- length(unique(long_data$Variable))


plots_per_page <- 9
n_pages <- ceiling(n_vars / plots_per_page)


for (i in 1:n_pages) {
  p <- ggplot(long_data, aes(x = Value)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap_paginate(~ Variable, scales = "free", ncol = 3, nrow = 3, page = i) +
    labs(title = paste("Numeric Variable Distributions – Page", i)) +
    theme_minimal()
  
  print(p)
}




```
Approx Normal, Symmetric: #Fld, #P, 1B_Pos, 3B_Pos, A, BA, BF, BK, C, DH, DP, H9, HitAllow, Rdrs/yr, RF, Rgood, SO/W, SS, W, WHIP
Approx Normal, right skew: #Bat, 2B_Bat, 3B_Bat, AllP, CF, CS, E, ER, ERA, FIP, HBP_Bat,HR_Bat, HR9,OBP, R/G, SP, TB, WP
Approx Normal, left skew: AB, BatAge, BB_Bat, Ch, DefEff, LOB_Bat, LOB_P,Rtot/yr, SF, SO_P, SO9, tSho
Not normal, Symmetric: 2B_Pos, CG_F, Fld%, H_Bats, HR_P, Rdrs, Rk
Not Normal, right skew:cSho, IBB_Bat, IBB_P, Non-p, OF(ALL), OPS
Not normal, left skew: BB_P, RP, Wins_Abv_Avg
not normal: BB9, ERA+, GDP, HBP_P,Inn, IP, LF, OPS+, PA, PAge, PH, PO, R_Bat_R_P, RA/G_P, RBI, Rtot, SB, SH, SLG, SO_Bat, SV


```{r}


numeric_data <- train_data %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")
#cor_matrix
high_corr <- as.data.frame(as.table(cor_matrix)) %>%
  rename(Var1 = Var1, Var2 = Var2, correlation = Freq) %>%
  filter(Var1 != Var2) %>%                        
  mutate(pair = paste(pmin(as.character(Var1), as.character(Var2)),
                      pmax(as.character(Var1), as.character(Var2)), sep = " - ")) %>%
  distinct(pair, .keep_all = TRUE) %>%            
  filter(abs(correlation) > 0.7) %>%              #
  arrange(desc(abs(correlation)))

high_corr


options(repr.plot.width = 15, repr.plot.height = 15)
corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.cex = 0.8)

```

There are too many variables to read on a standard correlation matrix. Instead I used chatgpt to create a table listing every pair with correlation of coefficient of 0.7 and higher. There is a lot of interaction between certain variables, as some variables might be a formula or a version of another variable(for example: Rtot and Rtot/yr which is just a yearly standardized version of Rtot or H9 which is hiter per 9 innings or hit allow which is total hits allowed). This table basically ask us to choose one variable or another as they are statistically and contextually the same.


```{r}


long_data <- train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(cols = -all_of(y_var), names_to = "Variable", values_to = "Value")


n_vars <- length(unique(long_data$Variable))


plots_per_page <- 9
n_pages <- ceiling(n_vars / plots_per_page)

for (i in 1:n_pages) {
  p <- ggplot(long_data, aes(x = Value, y = .data[[y_var]])) +
    geom_point(alpha = 0.6, color = "steelblue", size = 2) +
    facet_wrap_paginate(~ Variable, scales = "free_x", ncol = 3, nrow = 3, page = i) +
    labs(
      title = paste("Scatterplots of", y_var, "vs. Numeric Variables – Page", i),
      x = "Predictor Value",
      y = y_var
    ) +
    theme_minimal(base_size = 12)
  
  print(p)
}

```

Positive Relationship~
Strong linear:DefEff, Non-P, OPS+, SV, Wins_Abv_Avg
Moderate Linear:BB_Bat, DH, ERA+, HR_Bat, OF(ALL), OPS, R/G, R_Bat, RBI, SLG, SO_P, SO9, TB, tSho
Weak Linear: #Fld, 1b_pos, 3b_pos, BA, C, Fld%, LF, OBP, PA, PH, RF, SP

Negative Relationship~
Strong Linear: H9, HitAllow, Rk, WHIP
Moderate Linear: ER, ERA, LOB_P,
Weak Linear:BB_P, BB9, BF, E, FIP, HR_P, IBB_P, Inn, SH, SO/W

Not Pos or Neg but Linear:#Bat, #P, 2b_Bat, A, HBP_P, IBB_BAT, IP, SS, WP
No Relationship:2b_pos, 3b_BAt, AB, AllP, BatAge, BK, CF, CG_F, CS, cSho, GDP, H_Bats, HBP_Bat, HR9, PAge, PO, Rdrs, Rdrs/yr, Rgood, RP, Rtot, Rtot/yr, SB, SF, SO_Bat

(Quadratic or Parabola)
Stong Curvilinear:
Moderate CurvilinearL:LOB_BAT
Weak Curvilinear: Ch, DP

Log Relationship ~
Negative -
Strong: 
Moderate:R_P, RA/G
Weak:

Positive -
Strong:
Moderate:
Weak:

Nothing too suprising, only noticeable observation is that there are more, stronger or moderate linear relationships between defensive state(pitching and fielding) and wins than there are offensive stats and wins.
